---
title: "NHL Siblings"
format: revealjs
---

```{r}
library(tidyverse)
library(readr)
library(here)
```

```{r}
skaters <- read.csv(here("data/skaters.csv"))
```

```{r}
siblings <- read.csv(here("data/siblings.csv"))

siblings_full <- siblings |> pivot_longer(cols = c(sib1, sib2, sib3, sib4, sib5, sib6), names_to = "sib", values_to = "sibling", values_drop_na = TRUE)
```

```{r}
skaters_wsib <- left_join(skaters, siblings_full, by = join_by(player == sibling))
```

```{r}
has_sib <- skaters_wsib |> filter(!is.na(id))
no_sib <- skaters_wsib |> filter(is.na(id))
```


## Questions of Interest

Who are the best siblings, look at sibling pairs to see who has the most goals, assists, points, ect.

Are NHL players with siblings better or worse than players without siblings?

## Who are the best siblings

```{r}
g_a_sib <- has_sib |> group_by(family, id) |> summarise(tot_G = sum(G), tot_A = sum(A))

top_ten <- has_sib |> group_by(id, family) |> summarise(tot_G = sum(G), tot_A = sum(A)) |> arrange(desc(tot_A)) |> ungroup() |> slice(1:10)

ggplot(data = g_a_sib, aes(x = tot_G, y = tot_A)) +
    geom_point(aes(color = tot_G)) +
    geom_label(data = top_ten, aes(label = family)) +
    scale_color_viridis_c() +
    labs(title = "Siblings with the Most Goals and Assists") +
    theme_minimal(base_size = 15)
```

## Prominent Families

```{r}
gretzky <- skaters_wsib |> filter(family == "Gretzky") |> mutate(tot_G = G, tot_A = A, tot_GP = GP, tot_PIM = PIM)
sutter <- skaters_wsib |> filter(family == "Sutter") |> mutate(tot_G = G, tot_A = A, tot_GP = GP, tot_PIM = PIM)

ggplot(data = g_a_sib, aes(x = tot_G, y = tot_A)) +
    geom_point(aes(color = tot_G)) +
    geom_label(data = gretzky, aes(label = player)) +
    geom_label(data = sutter, aes(label = player)) +
    scale_color_viridis_c() +
    theme_minimal(base_size = 15) +
    labs(title = "Gretzky and Sutter Siblings")
```

## Siblings vs Non-Siblings Total

```{r}
sib_comp <- skaters_wsib |> mutate(sib_ind = if_else(!is.na(id), "Sib", "No Sib")) |>
    group_by(sib_ind) |> filter(!is.na(PIM)) |> filter(!is.na(plus_minus)) |>
    summarise(size = n(),
    tot_g = sum(G), 
    tot_a = sum(A), 
    tot_pts = sum(PTS),
    tot_pm = sum(plus_minus)
    ) |>
        pivot_longer(cols = c(size, tot_g, tot_a, tot_pts, tot_pm))
```

```{r}
ggplot(data = sib_comp, aes(x = name, y = value, fill = sib_ind)) +
    geom_col(position = "dodge") +
    labs(caption = "~7000 without siblings vs ~600 with sibling") +
    theme_minimal(base_size = 15)
```

## Siblings vs Non-Siblings per Player

```{r}
sib_comp_per_player <- skaters_wsib |> mutate(sib_ind = if_else(!is.na(id), "Sib", "No Sib")) |>
    group_by(sib_ind) |> filter(!is.na(PIM)) |> filter(!is.na(plus_minus)) |> 
    summarise(GP_per_player = sum(GP)/n(),
    g_per_player = sum(G)/n(), 
    a_per_player = sum(A)/n(), 
    pts_per_player = sum(PTS)/n(),
    PIM_per_player = sum(PIM)/n(),
    pm_per_player = sum(plus_minus)/n()
    ) |>
        pivot_longer(cols = c(GP_per_player, g_per_player, a_per_player, pts_per_player, PIM_per_player, pm_per_player))
```

```{r}
ggplot(data = sib_comp_per_player, aes(x = name, y = value, fill = sib_ind)) +
    geom_col(position = "dodge") +
    labs(title = "Stats per Player, Sibling vs. No Sibling") +
    theme_minimal(base_size = 15)

```

## Siblings vs Non-Siblings per Game

```{r}
sib_comp_per_game <- skaters_wsib |> mutate(sib_ind = if_else(!is.na(id), "Sib", "No Sib")) |>
    group_by(sib_ind) |> filter(!is.na(PIM)) |> filter(!is.na(plus_minus)) |> 
    summarise(g_per_game = sum(G)/sum(GP), 
    a_per_game = sum(A)/sum(GP), 
    pts_per_game = sum(PTS)/sum(GP),
    PIM_per_game = sum(PIM)/sum(GP),
    pm_per_game = sum(plus_minus)/sum(GP)
    ) |>
        pivot_longer(cols = c(g_per_game, a_per_game, pts_per_game, PIM_per_game, pm_per_game))
```

```{r}
ggplot(data = sib_comp_per_game, aes(x = name, y = value, fill = sib_ind)) +
    geom_col(position = "dodge") +
    labs(title = "Stats per game, Sibling vs. No Sibling") +
    theme_minimal(base_size = 15)
```


## New Questions

Noticing people have the same names, what pairs are most successful?

Are those pairs related or not?


## Same name

```{r}
same_name <- skaters |> group_by(player) |> mutate(size = n()) |> filter(size > 1) |>
    mutate(plus_minus = if_else(!is.na(plus_minus), plus_minus, 0, 0)) |>
    summarise(G = sum(G), A = sum(A), GP = sum(GP), PTS = sum(PTS), 
    plus_minus = sum(plus_minus), PIM = sum(PIM))

```

```{r}
same_name_topg <- same_name |> arrange(desc(G)) |> slice(1:10)
```

```{r}
ggplot(data = same_name, aes(x = GP, y = G)) +
    geom_point(aes(color = G)) +
    scale_color_viridis_c() +
    geom_label(data = same_name_topg, aes(label = player)) +
    labs(title = "Players with the same name") +
    theme_minimal()
```
